# Analysis of longitudinal data

_This week we apply the methods learned so far to longitudinal data, i.e. data collected over a longer timeperiod, with several observations per subject.: 

```{r, include=FALSE}
setwd("C:/Users/Ylva/Documents/GitHub/IODS-project")
library(dplyr)
library(tidyr)
library(ggplot2)
RATS <- read.table("data/rats.txt", sep = "\t")
BPRS <- read.table("data/bprs.txt", sep = "\t")

#Included this because for some reason these were returned as int
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

RATS$Group <- factor(RATS$Group)
RATS$ID <- factor(RATS$ID)
```


## Case 1: RATS!

### Overview 

The data used in this exercise, based on data by Kimmo Vehkalahti found [here](https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt), is from a nutrition study on three groups of rats, each rat's weight being recorded approximately once a week for 9 weeks. That is, the data is longitudinal data tracking the individuals over a time period. The goal is to compare the trends of each group of rats. The data has been wrangled to a long (as opposed to wide) format. 

Simply put, the variables are as follows: 

```{r echo = F}
head(RATS)
```

For a graphical overview, these are the trends over the 9-week time period across the three groups: 

```{r echo = F}
p1 <- ggplot(RATS, aes(x = Time, y = Weight, linetype = ID))+
  geom_line() + scale_linetype_manual(values = rep(1:10, times=4))+
  facet_grid(. ~ Group, labeller = label_both)+
  theme_bw() + theme(legend.position = "none")+
  theme(panel.grid.minor.y = element_blank()) +
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight))) + 
  ggtitle("Weights over time")
p1
```

The rats in the first group weigh noticeably less than the other groups, whereas the rats in group 2 seem to at the end of the experiment weigh closer to what the rats in group 3 did at the beginning. Group 2 also includes what seems to be an outlier. With standardised weights, the data appears as follows, allowing for a closer tracking of individuals: 

```{r echo = F}
# Standardise the scores:
RATSL <- RATS %>%
  group_by(Time) %>%
  mutate( stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
p1 <- ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() + scale_linetype_manual(values = rep(1:10, times=4))+
  facet_grid(. ~ Group, labeller = label_both)+
  theme_bw() + theme(legend.position = "none")+
  theme(panel.grid.minor.y = element_blank())+
  scale_y_continuous(name = "standardised weight") + 
  ggtitle("Standardised weights over time")
p1
```

### Summary measure

To see the forest for the trees - or the groups for the individuals - it is helpful to plot the data in a way that shows the groups, not individuals. 

```{r echo = F}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))+
  geom_boxplot(position = position_dodge()) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "right") +
  scale_x_discrete(name = "time") +
  scale_fill_grey(start = 0.5, end = 1) +
  ggtitle("Boxplots of groups")
p1
```

We shall now plot the summary measures, i.e. single values that represent all the observations of an individual. For this we use only the observations after the first measurement (Day 1). 

```{r echo = F}
# Make a summary data of the post treatment weeks (1-8)
RATS8S <- RATS %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

p1 <- ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot()+
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")+
  scale_y_continuous(name = "mean weight in g (since day 8)") +ggtitle("Summary measure of Groups")
p1
```
There are outliers in all Groups, so to remove bias we remove the outliers, i.e. the highest value from group 2 and the lowest values from Group 1 and Group 3. We repeat the plot without these outliers: 

```{r echo = F}
# Find outlies in Groups 1 & 2
max <- max(RATS8S$mean)
min <- min(RATS8S$mean)

# Find the outlier in Group 3
Group3 <- subset(RATS8S, Group == "3")
min3 <- min(Group3$mean)

RATS8S1 <- RATS8S %>%
  filter(mean < max & mean > min & mean != min3)

p2 <- ggplot(RATS8S1, aes(x = Group, y = mean)) +
  geom_boxplot()+
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")+
  scale_y_continuous(name = "mean weight in g (since day 8)") +ggtitle("Summary measure without outliers")
p2
```
Now the groups appear more obviously different. Group 1 is noticeably lighter in weight than the two other groups, and the weights of the rats in Group 3 are very close to each other. We shall verify the difference of Group 2 and 3 using Student's t-test. 

```{r echo = F}
# Without the outlier, apply Student's t-test, two-sided:
Groups12 <- subset(RATS8S1, Group != "1")
t.test(mean ~ Group, data = Groups12, var.equal = TRUE)
```
The p-value being 5.32e-05 confirms that these Groups indeed differ from each other in a statistically significant way. 


## Case 2: A Brief Psychiatric Rating Scale (BPRS)

### Overview 

BPRS is a measure on behaviour and psychiatric symptoms ([Wikipedia](https://en.wikipedia.org/wiki/Brief_Psychiatric_Rating_Scale)). In this dataset from [here](https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt) male subjects from two treatment groups have been tested and rated on the BPRS in one initial test (no. 0) plus eight weeks. This too, is a longitudinal dataset that has been wrangled into long form. The data looks something like this: 

```{r echo = F}
head(BPRS)
```

```{r}
ggplot(BPRS, aes(x=week, y=bprs, linetype=subject, fill = treatment)) + 
  geom_line() + 
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Week no.") +
  scale_y_continuous(name = "BPRS")+ 
  ggtitle("BPRS over time across 2 treatment groups") +
  theme_bw() + theme(legend.position = "none")
```
```{r}
ggplot(BPRS, aes(x = week, y = bprs, group = subject, color = treatment)) +
  geom_text(aes(label = treatment), position = position_dodge(width = 0.3)) +
  theme_bw() + ggtitle("BPRS of individuals in treatment groups")
```


In the first treatment group, the BPRS decreases over time for most individuals whereas in the second group the variation continues over the course of the testing. 

### Linear models

We shall fit a linear regression model for the data using the variables _treatment_ and _week_ as explanatory variables, and _bprs_ as the target variable. 

```{r echo = F}
BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRS)
summary(BPRSL_reg)
```

Based on this model, the treatment is not a statistically significant variable. However, it should be noted that for longitudinal data, a linear model is quite naive - even false - because it does not track individuals over time, i.e. it treats observations over time as independent of each other - which does not correspond to the reality of the situation. Thus, using the same variables as above, we must also fit a _random intercepts model_, which does take in account the longitudinal aspect of the data and assumes heterogeneity in intercepts. 

```{r echo = F}
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)
summary(BPRS_ref)
```

Below is a random slope model for the data, which assumes heterogeneity in slopes. As well as an analysis of variance (ANOVA) test on the above _random intercept_ and the below _random slope model_.  
```{r echo = F}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

writeLines("\n\nANOVA test\n")
# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

The ANOVA results indicate a fairly good fit of the models (p<0.05), i.e. the models are statistically significantly different from each other. We shall thus fit a _random intercept and slope model_. Instead of being distinguised, we allow for week*treatment interaction. We conduct an ANOVA to compare this with the above models. 

```{r echo = F}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week|subject), data = BPRS, REML = F)
summary(BPRS_ref2)

writeLines("\n\nANOVA test with random intercept model\n")
anova(BPRS_ref2, BPRS_ref)

writeLines("\n\nANOVA test with random slope model\n")
anova(BPRS_ref2, BPRS_ref1)
```

Evidently the difference to the _random intercept model_ is statistically significant, but the difference to the _random slope model_ less so (yet still p<0.1).

Finally, we shall draw a plot that shows the fitted values of the _random intercept and slope model_. 

```{r echo = F}
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRS <- BPRS %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(BPRS, aes(x = week, y = Fitted, linetype = subject, fill = treatment)) + 
  geom_line() + 
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Week no.") +
  scale_y_continuous(name = "BPRS")+ 
  ggtitle("BPRS with Fitted values") +
  theme_bw() + theme(legend.position = "none")
```

The difference is minimal even looking at the visualisation, and indeed [Vehkalahti & Everitt (2019: Chapter 8)](https://www.crcpress.com/Multivariate-Analysis-for-the-Behavioral-Sciences-Second-Edition/Vehkalahti-Everitt/p/book/9780815385158) confirm that the two treatment groups are not statistically significant. 

